---
title: "Assignment 6: Data Scraping"
author: "Tracy Che"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 

library(tidyverse)
library(rvest)
library(lubridate)
library(here)
library(cowplot)
library(RColorBrewer)

getwd()

here::i_am(
  "Assignments/TracyChe_A06_DataScraping.Rmd"
)

here()

mytheme <- theme_gray() +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "top"
  )

theme_set(mytheme)


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 


the_url <- paste0(
  "https://www.ncwater.org/wudc/app/lwsp/report.php?",
  "pwsid=03-32-010&year=2024"
)

the_website <- read_html(the_url)


```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

#name_css <- "div+ table tr:nth-child(1) td:nth-child(2)"
#pwsid_css <- "td tr:nth-child(1) td:nth-child(5)"
#ownership_css <- "div+ table tr:nth-child(2) td:nth-child(4)"
#maxday_css <- c("40.2400", "37.3800", "37.9100", "33.8100", "36.8400", "36.9000",
                #"36.5300", "38.0400", "33.8000", "37.9500", "35.2000", "32.2000")

the_name <- the_website %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text(trim = TRUE)

the_pwsid <- the_website %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text(trim = TRUE)

the_ownership <- the_website %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text(trim = TRUE)

the_maxday <- the_website %>%
  html_nodes('th~ td+ td') %>%
  html_text(trim = TRUE)

the_name
the_pwsid
the_ownership
the_maxday

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 


Durham2024 <- data.frame(
  System = rep(the_name, 12),
  PWSID = rep(the_pwsid, 12),
  Ownership = rep(the_ownership, 12),
  MaxDay = as.numeric(the_maxday),
  Month = c(
    "Jan", "May", "Sep",
    "Feb", "Jun", "Oct",
    "Mar", "Jul", "Nov",
    "Apr", "Aug", "Dec"
  )
) %>%
  mutate(
    Date = mdy(paste0(Month, "-1-2024"))
  ) %>%
  arrange(Date)

Durham2024


#5 

Durham2024 %>%
  ggplot(aes(x = Date, y = MaxDay)) +
  geom_line() +
  geom_point() +
  scale_x_date(
    breaks = "1 month",
    date_labels = "%b"
  ) +
  labs(
    title = "Durham Maximum Daily Water Use in 2024",
    x = "Month",
    y = "Maximum Day Use (MGD)"
  )

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}


#6

scrape.it <- function(PWSID, year){

  # Create the URL
  the_url <- paste0(
    "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
    PWSID,
    "&year=",
    year
  )
  # Read the website
  the_website <- read_html(the_url)

  # Scrape the data
  the_name <- the_website %>%
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
    html_text(trim = TRUE)

  the_ownership <- the_website %>%
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
    html_text(trim = TRUE)

  the_maxday <- the_website %>%
    html_nodes("th~ td+ td") %>%
    html_text(trim = TRUE)

  # Create dataframe
  the_df <- data.frame(
    System = rep(the_name, 12),
    PWSID = rep(PWSID, 12),
    Ownership = rep(the_ownership, 12),
    MaxDay = as.numeric(the_maxday),
    Month = c(
      "Jan", "May", "Sep",
      "Feb", "Jun", "Oct",
      "Mar", "Jul", "Nov",
      "Apr", "Aug", "Dec"
    ),
    Year = rep(year, 12)
  ) %>%
    mutate(
      Date = mdy(paste0(Month, "-1-", Year))
    ) %>%
    arrange(Date)

  # Return dataframe
  return(the_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 

Durham2020 <- scrape.it("03-32-010", 2020)

glimpse(Durham2020)

Durham2020 %>%
  ggplot(aes(x = Date, y = MaxDay)) +
  geom_line() +
  geom_point() +
  scale_x_date(
    breaks = "1 month",
    date_labels = "%b"
  ) +
  labs(
    title = "Durham Maximum Daily Water Use in 2020",
    x = "Month",
    y = "Maximum Day Use (MGD)"
  )

```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 

Cary2020 <- scrape.it("03-92-020", 2020)

water2020 <- bind_rows(
  Durham2020,
  Cary2020
)

water2020 %>%
  ggplot(aes(x = Date, y = MaxDay, color = System)) +
  geom_line() +
  geom_point() +
  scale_x_date(
    breaks = "1 month",
    date_labels = "%b"
  ) +
  labs(
    title = "Durham and Cary Maximum Daily Water Use in 2020",
    x = "Month",
    y = "Maximum Day Use (MGD)",
    color = "Water System"
  )


```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
Cary.multiyear <- map2(
  rep("03-92-020", 6),
  2018:2023,
  scrape.it
) %>%
  bind_rows()

Cary.multiyear %>%
  ggplot(aes(x = Date, y = MaxDay)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  scale_x_date(
    breaks = "1 year",
    date_labels = "%Y"
  ) +
  labs(
    title = "Cary Maximum Daily Water Use, 2018–2023",
    x = "Year",
    y = "Maximum Day Use (MGD)"
  )

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Yes. Cary has a slight increasing trend of water usage over the years. Although there are seasonal variations each year, the smooth line shows slight increase from around 22 MGD to over 24 MGD.  
>
