---
title: "Assignment 4: Data Wrangling"
author: "Tracy Che"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this assignment, you will build a reproducible workflow that cleans four EPA
air-quality files, combines them, reshapes the result, and produces grouped
summaries. Date conversion is a brief review from Module 3; the emphasis is on
data wrangling.

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

## Set up your session 

### Questions 1–2

1a. Load the `tidyverse` and `here` packages into your session.

1b. Check your working directory.

1c. Read all four raw files associated with the EPA Air dataset. See the README
file for information about the variables and measurements.

2. Add the appropriate code to reveal the dimensions of the four datasets. 

```{r setup questions 1-2, message = FALSE}
#1a 
library(tidyverse)
library(here)


#1b 
getwd()
here::i_am("Assignments/TracyChe_A04_DataWrangling.Rmd")
here()

#1c 

pm25_2024 <- read.csv(
  here("Data/A4/EPAair_PM25_NC2024_raw.csv")
)

pm25_2025 <- read.csv(
  here("Data/A4/EPAair_PM25_NC2025_raw.csv")
)

o3_2024 <- read.csv(
  here("Data/A4/EPAair_O3_NC2024_raw.csv")
)

o3_2025 <- read.csv(
  here("Data/A4/EPAair_O3_NC2025_raw.csv")
)
#2 

dim(pm25_2024)
dim(pm25_2025)
dim(o3_2024)
dim(o3_2025)

```

> All four datasets should have the same number of columns (PM2.5 has one extra column),
but unique record counts (rows). Do your datasets follow this pattern?
> Yes. The two PM2.5 datasets have 22 columns, while the two ozone datasets have 21 columns.

## Wrangle individual datasets to create processed files.

### Questions 3–5

3. Change the Date columns to be date objects.

4. Select the following columns: DDate, Daily.AQI.Value, Local.Site.Name, 
AQS.Parameter.Description, County, Site.Latitude, Site.Longitude

5. For the PM2.5 datasets, fill all cells in AQS.Parameter.Description with "PM2.5" 
(all cells in this column should be identical).

```{r questions 3 thru 5}
#3

pm25_2024$Date <- mdy(pm25_2024$Date)
pm25_2025$Date <- mdy(pm25_2025$Date)

o3_2024$Date <- mdy(o3_2024$Date)
o3_2025$Date <- mdy(o3_2025$Date)

# check
class(pm25_2024$Date)

#4


pm25_2024 <- select(
  pm25_2024,
  Date,
  Daily.AQI.Value,
  Local.Site.Name,
  AQS.Parameter.Description,
  County,
  Site.Latitude,
  Site.Longitude
)

pm25_2025 <- select(
  pm25_2025,
  Date,
  Daily.AQI.Value,
  Local.Site.Name,
  AQS.Parameter.Description,
  County,
  Site.Latitude,
  Site.Longitude
)

o3_2024 <- select(
  o3_2024,
  Date,
  Daily.AQI.Value,
  Local.Site.Name,
  AQS.Parameter.Description,
  County,
  Site.Latitude,
  Site.Longitude
)

o3_2025 <- select(
  o3_2025,
  Date,
  Daily.AQI.Value,
  Local.Site.Name,
  AQS.Parameter.Description,
  County,
  Site.Latitude,
  Site.Longitude
)

#5

pm25_2024 <- mutate(
  pm25_2024,
  AQS.Parameter.Description = "PM2.5"
)

pm25_2025 <- mutate(
  pm25_2025,
  AQS.Parameter.Description = "PM2.5"
)

```

### Question 6

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r question 6}
#6 


write.csv(
  pm25_2024,
  row.names = FALSE,
  file = here(
    "Data/A4/Processed/EPAair_PM25_NC2024_processed.csv"
  )
)

write.csv(
  pm25_2025,
  row.names = FALSE,
  file = here(
    "Data/A4/Processed/EPAair_PM25_NC2025_processed.csv"
  )
)

write.csv(
  o3_2024,
  row.names = FALSE,
  file = here(
    "Data/A4/Processed/EPAair_O3_NC2024_processed.csv"
  )
)

write.csv(
  o3_2025,
  row.names = FALSE,
  file = here(
    "Data/A4/Processed/EPAair_O3_NC2025_processed.csv"
  )
)

```


## Combine datasets

### Question 7

7. Stack the four datasets with `bind_rows()`. Check that their column names are
identical first.

```{r question 7}
#7 

#checking column names
colnames(pm25_2024)
colnames(pm25_2025)
colnames(o3_2024)
colnames(o3_2025)

#stack the rows
EPA.air.all <- bind_rows(
  pm25_2024,
  pm25_2025,
  o3_2024,
  o3_2025
)

dim(EPA.air.all)


```

### Question 8

8. Use one pipeline (`%>%`) to create a daily site-level dataset that meets the
following conditions:

  * Include only sites that the four data frames have in common:  
    
  "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",  
  "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School" 
  
  (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Group by date, site name,
  AQS parameter, and county, then calculate the mean AQI value, latitude, and
  longitude. Handle missing values explicitly in each mean.
  
  * Add numeric `Month` and `Year` columns from `Date`. Check if results are integers.
  
  * Hint: the dimensions of this dataset should be 15,012 × 9.

```{r question 8}
#8 


EPA.air.daily <- 
  EPA.air.all %>%
  filter(
    Local.Site.Name %in% c(
      "Linville Falls",
      "Durham Armory",
      "Leggett",
      "Hattie Avenue",
      "Clemmons Middle",
      "Mendenhall School",
      "Frying Pan Mountain",
      "West Johnston Co.",
      "Garinger High School",
      "Castle Hayne",
      "Pitt Agri. Center",
      "Bryson City",
      "Millbrook School"
    )
  ) %>%
  group_by(
    Date,
    Local.Site.Name,
    AQS.Parameter.Description,
    County
  ) %>%
  summarise(
    Daily.AQI.Value = mean(
      Daily.AQI.Value,
      na.rm = TRUE
    ),
    Site.Latitude = mean(
      Site.Latitude,
      na.rm = TRUE
    ),
    Site.Longitude = mean(
      Site.Longitude,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  mutate(
    Month = month(Date),
    Year = year(Date)
  )

#check
dim(EPA.air.daily)

```

### Questions 9–10

9. Use `pivot_wider()` so that ozone and PM2.5 AQI values occupy separate
columns. Each location-date combination should occupy one row.

10. Call up the dimensions of your new tidy dataset.

```{r questions 9-10}
#9


EPA.air.wide <- 
  EPA.air.daily %>%
  pivot_wider(
    names_from = AQS.Parameter.Description,
    values_from = Daily.AQI.Value
  )

#10

dim(EPA.air.wide)

```

### Question 11

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC2425_Processed.csv"

```{r question 11}
#11

write.csv(
  EPA.air.wide,
  row.names = FALSE,
  file = here(
    "Data/A4/Processed/EPAair_O3_PM25_NC2425_Processed.csv"
  )
)

```

## Generate summary tables

### Questions 12–13

12. Create a monthly summary data frame grouped by site, month, and year.
Calculate mean ozone and PM2.5 AQI for each group, handling missing values
explicitly. Then use `drop_na()` to remove rows only when mean **ozone** is
missing. Mean PM2.5 may remain missing.

13. Call up the dimensions of the summary dataset.

```{r questions 12 and 13}
#12


EPA.air.monthly.all <- 
  EPA.air.wide %>%
  group_by(
    Local.Site.Name,
    Month,
    Year
  ) %>%
  summarise(
    mean_ozone = mean(
      Ozone,
      na.rm = TRUE
    ),
    mean_pm25 = mean(
      PM2.5,
      na.rm = TRUE
    ),
    .groups = "drop"
  )


EPA.air.monthly <- 
  EPA.air.monthly.all %>%
  drop_na(mean_ozone)



#13

dim(EPA.air.monthly)

```

### Question 14

14. Why is `drop_na(mean_ozone)` more appropriate here than `na.omit()`? Test
both approaches and compare the dimensions of the resulting summary data frames.

```{r}

EPA.air.monthly.drop <- 
  EPA.air.monthly.all %>%
  drop_na(mean_ozone)

EPA.air.monthly.omit <- na.omit(
  EPA.air.monthly.all
)

dim(EPA.air.monthly.drop)
dim(EPA.air.monthly.omit)

```

> Answer: 
> the 'drop' is more appropriate because it removes rows only when mean ozone is missing. It keeps rows with valid ozone values even when mean PM2.5 is missing. In contrast, 'omit' removes rows if any variable contains a missing value, so there is less rows left after 'omit'. 


